Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2021-06-24 paper/tidy-missing-data-paper.Rmd
add plotting functions and stuff

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
    includes:
      in_header: "preamble.tex"
bibliography: "tidy-missing-data-zotero.bib"
cls: jss.cls
---

@@

```{r libs}
library(tidyverse) 
library(gridExtra)
library(ggthemes)
library(png)
library(scales)
library(simputation)
library(naniar) 
library(knitr)
library(lubridate)
library(magick)
library(rpart)
library(visdat)
library(patchwork)
```

@@

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools suffer the same problems as modelling: they use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis. The complex, often multivariate nature of imputation methods also makes makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.

@@

\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix "\code{_imp}". So, "\code{Var1}" has a sibling column, "\code{Var1_imp}", with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in Section \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.

@@

\pkg{ggplot2} removes missing values with a warning (Figure \@ref(fig:warning)), and only incorporates missingness into visualizations when mapping a discrete variable with missings to a graph aesthetic. This has some limitations, shown in Figure \@ref(fig:gg-box-na), a boxplot visualization of school grade and test scores. If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed (Figure \@ref(fig:gg-box-na)).

@@

tibble::tibble(
  `Function` = c(
    "add_n_miss(data)",
    "add_prop_miss(data)",
    "add_miss_cluster(data)"
    ),
  `Adds column which:` = c(
    "contains the number missing values in a row",
    "contains the proportion of missing values in a row",
    "contains the missing value cluster"
  )) %>% 
    knitr::kable(caption = 'Overview of the "add" functions in naniar')

@@

```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in the airquality data. Missings invariables (A) and cases (B), and for a birds eye view with missingness as a heatmap in (C), and with clustering applied (D). There are missing values in Ozone and Solar.R, with Ozone having more missings. Not many cases have two missings. Most missingness is from cases with one missing value. The default output (C) and ordered by clustering on rows and columns (D). These overviews are made possible using the shadow matrix in long form. There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness.", fig.height = 7, fig.width = 6, out.width = "90%", fig.align="center"}

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into *nabular* form, then values are imputed below the range of data using \code{impute_below_all()}. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and "Ozone_NA" is mapped to the fill aesthetic. *Nabular* data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.

@@

It is useful to explore summaries and tabulations within groups of a dataset. \pkg{naniar} works with \pkg{dplyr}'s \code{group_by()} operator to produce grouped summaries, which work well with the "pipe" operator. The code and Table \@ref(tab:group-miss-var-summary) below show an example of missing data summaries for airquality, grouped by month.

@@

```{r data-setup, include = FALSE}

housing_raw <-
  readr::read_csv(here::here("data",
                             "melbourne_housing_raw.csv")) %>%
  janitor::clean_names() %>%
  dplyr::rename(region_name = regionname,
         property_count = propertycount) %>%
  dplyr::mutate(date = lubridate::dmy(date)) %>%
  dplyr::rename(latitude = lattitude,
                longitude = longtitude) %>%
  # let's create monthly quarters
  mutate(yr_qtr = as.ordered(lubridate::quarter(date, 
                                             with_year = TRUE,
                                             fiscal_start = 1))) %>%
  # drop price
  tidyr::drop_na(price) %>%
  # make price the log of price
    mutate(price = log10(price))

```{r data-clean, include = FALSE}

# compact down the seller levels
count_dat <- count(housing_raw,seller_g) %>% 
  arrange(-n) %>%
  mutate(cumu = cumsum(n),
         pct = cumu / nrow(housing_raw),
         pct_change = pct - lag(pct)) %>%
  tibble::rowid_to_column() %>%
  mutate(seller = case_when(
    pct < 0.52 ~ as.character(seller_g),
    pct > 0.52 & pct < 0.60 ~ "seller_g1",
    pct > 0.60 & pct < 0.65 ~ "seller_g2",
    pct > 0.65 & pct < 0.70 ~ "seller_g3",
    pct > 0.70 & pct < 0.75 ~ "seller_g4",
    pct > 0.75 & pct < 0.80 ~ "seller_g5",
    pct > 0.80 & pct < 0.85 ~ "seller_g6",
    pct > 0.85 & pct < 0.90 ~ "seller_g6",
    pct > 0.90 & pct < 0.95 ~ "seller_g6",
    pct > 0.95 ~ "seller_g5"
  ))

# drop these three variables, as we won't use them in analysis, they
# would need extensive recoding

housing <- housing_raw %>%
  mutate_if(is.character, as.factor) %>%
  mutate_at(vars(rooms,
                 bedroom2,
                 bathroom,
                 car),
            as.factor) %>%
  # recode / collapse factors with many levels and few obs
  mutate(bathroom = fct_other(bathroom,
                                drop = c("4", "5", "6", "7", "8", "9"),
                                other_level = "4+")) %>%
  mutate(bedroom2 = fct_other(bedroom2,
                               drop = c("5", "6", "7", "8", "9", "10", 
                                        "12", "16", "20"),
                               other_level = "5+")) %>%
  mutate(rooms = fct_other(rooms,
                             drop = c("6", "7", "8", "9", "10", 
                                      "12", "16"),
                             other_level = "6+")) %>%
  mutate(car = fct_other(car,
                           drop = c("4", "5", "6", "7", "8", "9", "10",
                                    "11", "18"),
                           other_level = "4+")) %>%
  left_join(select(count_dat, seller_g, seller), by = "seller_g")  %>%
  # drop seller_g
  select(-seller_g) %>%
  mutate(seller = as.factor(seller))

@@

Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values. The most missings are in "building area", followed by "year built", and "land size", with similar amounts of missingness in "Car", "bathroom", "bedroom2", "longitude", and "latitude." Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing. The variables "building area" and "year built" have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious. Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).

```{r housing-miss-case-var, fig.cap = "The amount of missings in variables (A) and cases (B) for Melbourne housing data. (A) Build area and year built have more than 50\\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings. (B) Cases are missing 5 - 50\\% of values. The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}
q1 <- gg_miss_var(housing, show_pct = TRUE) + labs(tag = "A")

@@

housing %>%
  # sample_frac(0.25) %>%
  vis_miss(cluster = TRUE,
           sort_miss = TRUE,
           show_perc_col = FALSE)

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
    affiliation2: Telethon Kids Institute, Perth Children's Hospital, Perth, Australia
  - name: Dianne Cook
    affiliation: Monash University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
  # For running headers, if needed
  short:     "Explore missingness with \\pkg{naniar}"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
keywords:
  # at least one keyword must be supplied
  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    keep_tex: yes
    citation_package: 'natbib'
    includes:
      in_header: "preamble.tex"
bibliography: "tidy-missing-data-zotero.bib"
cls: jss.cls
---

@@

```{r libs}
library(tidyverse) 
library(gridExtra)
library(ggthemes)
library(png)
library(scales)
library(simputation)
library(naniar) 
library(knitr)
library(lubridate)
library(magick)
library(rpart)
library(visdat)
library(patchwork)
library(withr)
library(here)
library(targets)
```

```{r tar-load}
with_dir(here(), {
  tar_load(c(housing, vis_miss_housing))
})
```

@@

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools suffer the same problems as modelling: they use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis. The complex, often multivariate nature of imputation methods also makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.

@@

\pkg{VIM} [@VIM] implements well-used imputation methods: K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allow for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix "\code{_imp}". So, "\code{Var1}" has a sibling column, "\code{Var1_imp}", with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in Section \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal component analysis.

@@

\pkg{ggplot2} removes missing values with a warning (Figure \@ref(fig:warning)), and only incorporates missingness into visualizations when mapping a discrete variable with missings to a graph aesthetic. This has some limitations, shown in Figure \@ref(fig:gg-box-na), a boxplot visualization of school grade and test scores. If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed.

@@

tibble::tibble(
  `Function` = c(
    "add_n_miss(data)",
    "add_any_miss(data)",
    "add_prop_miss(data)",
    "add_miss_cluster(data)"
    ),
  `Adds column which:` = c(
    "contains the number missing values in a row",
    "contains whether there are any missing values in a row",
    "contains the proportion of missing values in a row",
    "contains the missing value cluster"
  )) %>% 
    knitr::kable(caption = 'Overview of the "add" functions in naniar')

@@

```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in the airquality data. Missings in variables (A) and cases (B), and for an overview of all missingness as a heatmap in (C), and with clustering applied (D). There are missing values in Ozone and Solar.R, with Ozone having more missings. Not many cases have two missings. Most missingness is from cases with one missing value. The default output (C) and ordered by clustering on rows and columns (D). These overviews are made possible using the shadow matrix in long form. There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness.", fig.height = 7, fig.width = 6, out.width = "90%", fig.align="center"}

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into *nabular* form, then values are imputed below the range of data using \code{impute_below_all()}. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and "Ozone_NA" is mapped to the fill aesthetic. *Nabular* data facilitates adding counts of missingness to a histogram, allowing examination of a variable's distribution of values, and also the magnitude of missings.

@@

It is useful to explore summaries and tabulations within groups of a dataset. \pkg{naniar} works with \pkg{dplyr}'s \code{group_by()} operator to produce grouped summaries, which work well with the "pipe" operator. Table \@ref(tab:group-miss-var-summary) below shows an example of missing data summaries for airquality, grouped by month.

@@

@@

Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values. The most missings are in "building area", followed by "year built", and "land size", with similar amounts of missingness in "Car", "bathroom", "bedroom2", "longitude", and "latitude". Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing. The variables "building area" and "year built" have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious. Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).

```{r housing-miss-case-var, fig.cap = "The amount of missings in variables (A) and cases (B) for Melbourne housing data. (A) Build area and year built have more than 50\\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings. (B) Cases are missing 5 - 50\\% of values. The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}

q1 <- gg_miss_var(housing, show_pct = TRUE) + labs(tag = "A")

@@

vis_miss_housing
